Gusto: a toolkit for compatible finite element dynamical cores

### 2024-03-10 source/paper.tex
some edits

%% \usepackage commands included in the copernicus.cls:
%\usepackage[german, english]{babel}
%\usepackage{tabularx}
%\usepackage{cancel}
%\usepackage{multirow}
%\usepackage{supertabular}
%\usepackage{algorithmic}
%\usepackage{algorithm}
%\usepackage{amsthm}
%\usepackage{float}
%\usepackage{subfig}
%\usepackage{rotating}

@@

In this article we will describe the current capabilities of Gusto and
the design features that enable our flexible implementation whereby
different spatial and temporal discretisations can be applied to a
range of different equation sets relevant for weather and climate
modelling. In section \ref{sec: governing} we will outline three
different equation sets, along with their finite element
discretisations. Then in section \ref{sec: design} we will present the
code structure of Gusto followed by a detailed example (section
\ref{sec: FML}) of the Form Manipulation Language (FML) that facilitates
symbolic manipulation of the finite element forms within our
timestepping code. Section \ref{sec: results} contains a selection of
numerical results that showcases the range of discretisations
available within Gusto. We summarise our current status and future
directions in section \ref{sec: summary}.

\section{Governing Equations}
\label{sec: governing}
The aim of this section is to introduce three governing equation sets
commonly used in the development of weather and climate models, along
with enough detail about the finite element discretiations provided by
Gusto to demonstrate the similarities between them. These similarities
guide the design of the code structure described in the next section
and motivate the use of Form Manipulation Labelling (FML) that
underlies the flexibility of Gusto, a detailed example of which will
be given in section \ref{sec: FML}. The convention in Gusto is to
write the equations in residual form, meaning that all the terms
appear on the left hand side. We will follow this convention below in
contrast to many papers which present the equations with the pressure
gradient term and any forcing terms on the right hand side.

@@

Comparing the equation sets given in the previous sections, we see
several similarities. For example, the prognostic equation for the
velocity always contains a nonlinear transport term, a term due to
rotation and a `pressure' gradient term (in shallow water this is the
gradient of the depth but the term plays the same role in the dynamics
and is derived from the pressure gradient term in the full equations
(**cite Zeitlin book)). The other prognostic fields satisfy either an
advective or continuity form of the transport equation. These
similarities remain when we formulate the compatible finite element
discretisations, as we shall see below, and importantly this means
that the discretisation of these terms can be written in a general way
that means it can be applied to any of the equation sets.

@@

\begin{equation}
\int_\Omega\MM{w}\cdot\nabla p \diff x = -\int_\Omega\nabla\cdot\MM{w}p\diff x,
\end{equation}
where $\Omega$ denotes the domain and there are no surface terms from
the integration by parts because these vanish on interior element
boundaries since the normal component of $\MM{w}$ is continuous across
element faces and on exterior boundaries we set no normal flow
boundary conditions. However, the pressure gradient term in the
compressible Euler equations, $c_p\theta\nabla\Pi$, is nonlinear and
this introduces an additional complication. Integrating by parts is
still required since $\Pi$ is evaluated as a function of the density
$\rho \in L_2$ and so is discontinuous, but this means that we now
have a differential operator applied to $\theta \in \mathbb{V}_\theta$
which is discontinuous in the vertical. Following \ref{} we integrate
by parts in each element $e$, giving

\begin{equation}
  c_p\int_e\MM{w}\cdot\theta\nabla\Pi\diff x = -c_p\int_e\nabla\cdot(\MM{w}\theta)\Pi\diff x + c_p\int_{\partial e}\theta^-\MM{w}\cdot\MM{n}\Pi^- \diff S,
\end{equation}
where $\partial e$ denotes the element faces, $\MM{n}$ is the outward
pointing normal to the face and $\cdot^-$ denotes the value of a
variable on the interior side of the boundary. Summing over the domain
gives

\begin{equation}
  c_p\int_\Omega\MM{w}\cdot\theta\nabla\Pi\diff x = -c_p\int_\Omega\nabla\cdot(\MM{w}\theta)\Pi\diff x + c_p\int_{\Gamma_v}\jump{\theta\MM{w}\cdot\MM{n}}\avg{\Pi} \diff S,
\end{equation}
where $\jump{\cdot}$ and $\avg{\cdot}$.

@@

\begin{itemize}
\item include schematic
\item generic example - no FML
\item lead in to next section... facilitated by FML i.e. timesteppers don't care what equations they're given because forms are labelled
\end{itemize}

@@

\subsection{Shallow water simulations}
The standard suite of shallow water test cases originally proposed in
\citet{williamson1992standard} provides a common starting point for
testing new discretisations. This test suite is usually augmented by
the unstable barotropic jet described in \citet{galewsky2004initial}
(**and linear tests?)

>>>>

%% \usepackage commands included in the copernicus.cls:
%\usepackage[german, english]{babel}
%\usepackage{tabularx}
%\usepackage{cancel}
%\usepackage{multirow}
%\usepackage{supertabular}
\usepackage{algorithmic}
\usepackage{algorithm}
%\usepackage{amsthm}
%\usepackage{float}
%\usepackage{subfig}
%\usepackage{rotating}

@@

In this article we will describe the current capabilities of Gusto and
the design features that enable our flexible implementation whereby
different spatial and temporal discretisations can be applied to a
range of different equation sets relevant for weather and climate
modelling. In section \ref{sec: governing} we will outline three
different equation sets, along with their finite element
discretisations. Then in section \ref{sec: design} we will present the
code structure of Gusto followed by a detailed example (section
\ref{sec: FML}) of the Form Manipulation Language (FML) that
facilitates symbolic manipulation of the finite element forms within
our timestepping code. Section \ref{sec: results} contains a selection
of numerical results that showcases the range of discretisations
available within Gusto. **mention here new functionality facilitated
by the flexibility of Gusto? We summarise our current status and future
directions in section \ref{sec: summary}.

\section{Governing Equations}
\label{sec: governing}
The aim of this section is to introduce three governing equation sets
commonly used in the development of weather and climate models, along
with enough detail about the finite element discretiations provided by
Gusto to demonstrate the similarities between them. These similarities
guide the design of the code structure described in the next section
and motivate the use of Form Manipulation Language (FML) that
underlies the flexibility of Gusto, a detailed example of which will
be given in section \ref{sec: FML}. The convention in Gusto is to
write the equations in residual form, meaning that all the terms
appear on the left hand side. We will follow this convention below in
contrast to many papers which present the equations with the pressure
gradient term and any forcing terms on the right hand side.

@@

Comparing the equation sets given in the previous sections, we see
several similarities. For example, the prognostic equation for the
velocity always contains a nonlinear transport term, a term due to
rotation and a `pressure' gradient term (in shallow water this is the
gradient of the depth but the term plays the same role in the dynamics
and is derived from the pressure gradient term in the full equations
\citep{zeitlin2018geophysical}). The other prognostic fields satisfy
either an advective or continuity form of the transport
equation. These similarities remain when we formulate the compatible
finite element discretisations, as we shall see below, and importantly
this means that the discretisation of these terms can be written in a
general way that means it can be applied to any of the equation sets.

@@

\begin{equation}
\int_\Omega\MM{w}\cdot\nabla p \diff x = -\int_\Omega\nabla\cdot\MM{w}p\diff x,
\end{equation}
where $\Omega$ denotes the domain and there are no surface terms from
the integration by parts because these vanish on interior element
boundaries since the normal component of $\MM{w}$ is continuous across
element facets and on exterior boundaries we set no normal flow
boundary conditions. However, the pressure gradient term in the
compressible Euler equations, $c_p\theta\nabla\Pi$, is nonlinear and
this introduces an additional complication. Integrating by parts is
still required since $\Pi$ is evaluated as a function of the density
$\rho \in L_2$ and so is discontinuous, but this means that we now
have a differential operator applied to $\theta \in \mathbb{V}_\theta$
which is discontinuous in the vertical. Following \ref{} we integrate
by parts in each element $e$, giving

\begin{equation}
  c_p\int_e\MM{w}\cdot\theta\nabla\Pi\diff x = -c_p\int_e\nabla\cdot(\MM{w}\theta)\Pi\diff x + c_p\int_{\partial e}\hat{\theta}\MM{w}\cdot\MM{n}\hat{\Pi}\diff S,
\end{equation}
where $\partial e$ denotes the element facets, $\MM{n}$ is the outward
pointing normal to the facet and $\hat{q}$ denotes the value of a
variable $q$ on the facet, which for discontinuous variables does not
have a unique definition. Summing over the domain gives

\begin{equation}
  c_p\int_\Omega\MM{w}\cdot\theta\nabla\Pi\diff x = -c_p\int_\Omega\nabla\cdot(\MM{w}\theta)\Pi\diff x + c_p\int_{\Gamma_v}\jump{\theta\MM{w}\cdot\MM{n}}\avg{\Pi} \diff S,
\end{equation}
where $\Gamma_v$ denotes the vertical element facets, across which
$\theta$ is discontinuous. Here we have chosen, as in \ref{}, to take
the average value of $\Pi$ across the facets, $\avg{\Pi} =
(\Pi^++\Pi^-)/2$, where the $+$ and $-$ indicate each side fo the
facet. The brackets $\jump{q} = q^+-q^-$ denote the jump in a value
across the facet and the integral over the horizontal facets vanishes
since both $\theta$ and $\MM{w}\cdot\MM{n}$ are continuous across
these facets so the jump is zero.

@@

\begin{algorithm}
  \caption{Pseudocode for SIQN timestep loop}\label{alg:SIQN}
  \begin{algorithmic}
    \STATE
  \end{algorithmic}
\end{algorithm}

\begin{itemize}
\item generic example(s) - no FML, just pseudocode for basic
  timestepper, SIQN and split physics
\item lead in to next section... facilitated by FML i.e. timesteppers don't care what equations they're given because forms are labelled
\end{itemize}

@@

\subsection{Terminator `toy' chemistry}
This test combines transport and a `toy' chemistry model to
demonstrate physics-dynamics coupling in Gusto (where here chemistry
is the `physics').

\subsection{Shallow water}
The standard suite of shallow water test cases originally proposed in
\citet{williamson1992standard} provides a common starting point for
testing new discretisations. This test suite is usually augmented by
the unstable barotropic jet described in \citet{galewsky2004initial}
